---
title: "Calibrarion of Ordinal Posterior Predictions"
date: "2023-02-04"
date-modified: last-modified
author: "Teemu Säilynoja"
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    code-line-numbers: true
    default-image-extension: png
    fig-format: png
    embed-resources: true
knitr:
  opts_chunk:
    fig.path: ../../images/
    dev.args:
      bg: transparent
execute:
  cache: true
---

This notebook reproduces and expands on the examples of predictive checks for ordinal data, shown in
Section 6 of the article.

```{r}
#| label: options
#| code-summary: "Imports & options"
#| output: false
#| message: false

library("bayesplot")
library("cmdstanr")
library("patchwork")
library("ggplot2")
library("khroma")
library("dplyr")


# Source for the modified reliability plot
source("code/R/helpers.R")
source("code/R/pava_plots.R")

theme_set(ppc_paper_theme())
bayesplot_theme_set(ppc_paper_theme())

SEED <- 236543
set.seed(SEED)
SAVE_FITS <- TRUE
```


In assessing the calibration, we use the ordinal nature of the data and use the cumulative posterior
predictive mass function.

## Data set

```{r}
#| label: data_generation
#| code-summary: Data generation
K <- 5
N <- 1500
sigma <- .5
c <- sample(1:K, N, replace = T)
x <- rnorm(N, c, sigma)
standata_gmm <- list(
  K = K,
  N = N,
  x = x,
  y = c,
  sigma = sigma
)

```

We generate data by drawing `r N` observations from a mixture of `r K` Gaussians with
means `r paste(1:(K-1), sep = ", ")`, and `r K`, and standard deviation $0.5$. 

\begin{align}
x_n &\sim \mathcal N\!\!\left(k, 0.5^2\right), &\text{for } n \in\{1,\dots,N\}\\
k &\sim \text{Categorical}(\theta_k),&\\
\theta_k &= \frac 1 K, &\text{for } k \in \{1, \dots, K\}.
\end{align}
```{r}
#| label: 4_3_ordinal_data_densities
#| dev.args :
#|   bg: transparent
data.frame(
  x = rep(seq(min(standata_gmm$x), max((standata_gmm$x)), .01), K),
  d = 1:K |> sapply(\(c) dnorm(seq(min(standata_gmm$x), max((standata_gmm$x)), .01), c, sigma)) |> c(),
  c = rep(1:K, each = length(seq(min(standata_gmm$x), max((standata_gmm$x)), .01)))
) |>
  group_by(x) |>
  mutate(percentage = d / sum(d)) |>
  ggplot() +
  geom_line(
    aes(
      x = x,
      y = percentage,
      color = as.factor(c),
      group = c
    ),
    linewidth = 2,
    key_glyph = "rect"
  ) +
  scale_color_manual(
    aesthetics = c("color"),
    values = unname(paper_colors[1:6])
  ) +
  labs(color = "Value", x = expression("x"), y = "Proportion") +
  coord_equal(ratio = 2, xlim = range(standata_gmm$x), ylim = c(0, 1), expand = FALSE) +
  theme(legend.position = "top")
```

## Model

We fit two models to the data, both structured to first normalize the data and
then fit a Gaussian mixture model (GMM) with K = `r K`` mixture components. The first model replicates a simple implementation error where standard deviation of the mixture components was incorrectly scaled.
```{r}
#| label: model_code
#| code-summary: Read model code

gmm <- cmdstan_model("code/stan/gmm_classifier.stan")
gmm
```

```{r}
#| warning: false
#| label: fit_models
#| code-summary: run CmdStanR
fit_1 <- gmm$sample(
  data = c(standata_gmm, list(correct_sigma = 0)),
  parallel_chains = 4,
  refresh = 0,
  seed = SEED,
  show_messages = FALSE
)

fit_2 <- gmm$sample(
  data = c(standata_gmm, list(correct_sigma = 1)),
  parallel_chains = 4,
  refresh = 0,
  seed = SEED,
  show_messages = FALSE
)
```

```{r}
#| label: classification probabilities
p_1 <- matrix(colMeans(fit_1$draws(variables = "ppm", format = "matrix")), ncol = K)

p_2 <- matrix(colMeans(fit_2$draws(variables = "ppm", format = "matrix")), ncol = K)
```

## Predictive checks

First, we use a bar graph to visualize the relative frequencies of the groups in the observation and the predictive draws. We don't see much difference between the models, and all of the observations seem to fall within the credible intervals of the predictions of both models.

```{r}
#| label: 4_3_ordinal_ppc_bars
p2 <- ppc_bars(
  y = as.numeric(c),
  yrep = fit_2$draws(variables = "yrep", format = "matrix")
) +
  ggtitle("Model 2") +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank()
  )

p1 <- ppc_bars(
  y = as.numeric(c),
  yrep = fit_1$draws(variables = "yrep", format = "matrix")
) +
  ggtitle("Model 1")

(p1 + p2) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

When using the PAV-adjusted calibration plots, we see that the first model, with the initial implementation error is under confident (S-shaped calibration curves) in its predictions, and assigns probabilities too close to 0.5 to many cases that the other model is better able to separate towards the observed outcomes.

```{r}
#| label: 4_3_ordinal_reliability_plots_1
ppc_calibration_pava(
  y = as.numeric(c <= 1),
  p = p_1[, 1],
  quantiles = 100,
  dot_scale = .7,
  fill_alpha = .3,
  cep_line_color = paper_colors["orange"]
)
ppc_calibration_pava(
  y = as.numeric(c <= 2),
  p = pmin(1, rowSums(p_1[, 1:2])),
  quantiles = 100,
  dot_scale = .4,
  fill_alpha = .3,
  cep_line_color = paper_colors["orange"]
)
ppc_calibration_pava(
  y = as.numeric(c <= 3),
  p = pmin(1, rowSums(p_1[, 1:3])),
  quantiles = 100,
  dot_scale = .4,
  fill_alpha = .3,
  cep_line_color = paper_colors["orange"]
)
ppc_calibration_pava(
  y = as.numeric(c <= 4),
  p = pmin(1, rowSums(p_1[, 1:4])),
  quantiles = 100,
  dot_scale = .7,
  fill_alpha = .3,
  cep_line_color = paper_colors["orange"]
)
```


```{r}
#| label: 4_3_ordinal_reliability_plots_2
ppc_calibration_pava(
  y = as.numeric(c <= 1),
  p = p_2[, 1],
  quantiles = 100,
  dot_scale = .75,
  fill_alpha = .3,
  cep_line_color = paper_colors["orange"]
)
ppc_calibration_pava(
  y = as.numeric(c <= 2),
  p = pmin(1, rowSums(p_2[, 1:2])),
  quantiles = 100,
  dot_scale = .7,
  fill_alpha = .3,
  cep_line_color = paper_colors["orange"]
)
ppc_calibration_pava(
  y = as.numeric(c <= 3),
  p = pmin(1, rowSums(p_2[, 1:3])),
  quantiles = 100,
  dot_scale = .7,
  fill_alpha = .3,
  cep_line_color = paper_colors["orange"]
)
ppc_calibration_pava(
  y = as.numeric(c <= 4),
  p = pmin(1, rowSums(p_2[, 1:4])),
  quantiles = 100,
  dot_scale = .85,
  fill_alpha = .3,
  cep_line_color = paper_colors["orange"]
)
```




